---
title: "4) DE by condition"
author:
- Kane Toh^[Genomics and Data Analytics Core (GeDaC), Cancer Science Institute of Singapore, National University of Singapore; kanetoh@nus.edu.sg]
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: yes
  bookdown::html_document2:
    code_folding: show
    df_print: paged
    theme: cerulean
    highlight: tango
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
urlcolor: blue
---

```{r chunk_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, 
                      message=FALSE,
                      echo=TRUE,
                      include=TRUE)
```

```{r library}
library(Seurat)
library(dplyr)
library(tidyr)
library(ggplot2)
library(clusterProfiler)
library(msigdbr)
library(rjson)
```

```{r globalVar}
# Loads in custom functions that I've written to assist with plotting/saving tasks
source("helper.R")

# Path to saving the results fo the analysis
res_path <- "../results/04_DE_condition/"
if (!dir.exists(res_path)){
  
  dir.create(res_path)
}else{
  print("Results directory exists")
}
# Data path specifes where all the outputs from the parseBio split-pipe pipeline is stored.
data_path <- "../data/"
if (!dir.exists(data_path)){
  warning("Directory does not exist: ", data_path)
}else{
  print("Data directory exists")
}

# RDS path specifies where all the intermediate generated R objects will be saved.
rds_path <- "../rds/"
if (!dir.exists(rds_path)){
  warning("Directory does not exist: ", rds_path)
}else{
  print("RDS directory exists")
}
```

# Load seurat object

```{r load-qced-object}
mbladder <-  ReadObject(rds_path, "03_seuratMerged_annotated")
# Rerun scaledata again on all genes (instead of just top 2000 variable genes by default in seurat)
mbladder <- ScaleData(mbladder, features=rownames(mbladder))
```
# Set identity to condition for subsequent comparison

```{r}
Idents(mbladder) <- mbladder@meta.data$Condition
```


# Find DE genes between condition for all cells.

```{r}
#condition_markers <- FindMarkers(mbladder, ident.1 = "KO", ident.2="WT",
#                                logfc.threshold = -Inf, min.pct = -Inf, min.diff.pct = -Inf)
marker_genes <- FindAllMarkers(mbladder, only.pos=TRUE)
marker_genes_sig_condition <- marker_genes %>%
  dplyr::filter(p_val_adj<0.05)
marker_genes_sig_condition_top_30 <- marker_genes_sig_condition %>%
  dplyr::group_by(cluster) %>%
  dplyr::arrange(desc(avg_log2FC)) %>%
  dplyr::slice_head(n=30) %>%
  pull()
marker_genes_sig_condition %>% write.csv(file=paste0(res_path, "marker_genes_byCondition_onlyPos.csv"))
marker_genes_sig_condition_genelist <- marker_genes_sig_condition %>%
  dplyr::select(gene) %>%
  pull()
```

# Plotting heatmap of all cells grouped by condition
```{r}
heatmap_plot <- DoHeatmap(mbladder, features = marker_genes_sig_condition_top_30,
                          label=FALSE)
SaveFigure(res_path = res_path,
           plots = heatmap_plot,
           name = "heatmap_conditions", 
           width = 15, 
           height = 8)
```
# KEGG analysis of KO vs WT

```{r}
# First, obtian the genes that are statistically significantly enrihed in KO vs WT
KO_enriched_genes <- marker_genes_sig_condition %>%
  dplyr::filter(cluster == "KO") %>%
  pull()

WT_enriched_genes <- marker_genes_sig_condition %>%
  dplyr::filter(cluster == "WT") %>%
  pull()

# Next, convert gene symbols to entrez ID by querying BioMart to convert symbols to Entrez IDs
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
# 
entrez_KO_genes <- getBM(attributes = c('entrezgene_id'),
                     filters = 'mgi_symbol',
                     values = KO_enriched_genes,
                     mart = ensembl) %>% pull()

entrez_WT_genes <- getBM(attributes = c('entrezgene_id'),
                     filters = 'mgi_symbol',
                     values = WT_enriched_genes,
                     mart = ensembl) %>% pull()
# Perform KEGG Overrepresentation analysis
KO_genes_kegg <- enrichKEGG(gene = entrez_KO_genes,
                 organism     = 'mmu',
                 pvalueCutoff = 0.05)
WT_genes_kegg <- enrichKEGG(gene = entrez_WT_genes,
                 organism     = 'mmu',
                 pvalueCutoff = 0.05)
# Plot
dotplot_KEGG_KO <- dotplot(KO_genes_kegg, showCategory=10, font.size=12) + ggtitle("KEGG Pathway enrichment for KO-upregulated genes")
dotplot_KEGG_KO
SaveFigure(res_path = res_path,
           plots = dotplot_KEGG_KO,
           name = "dotplot_KEGG_KO", 
           width = 15, 
           height = 8)

dotplot_KEGG_WT <- dotplot(WT_genes_kegg, showCategory=10, font.size=12) + ggtitle("KEGG Pathway enrichment for WT-upregulated genes")
SaveFigure(res_path = res_path,
           plots = dotplot_KEGG_WT,
           name = "dotplot_KEGG_WT", 
           width = 15, 
           height = 12)
```

# MSigDB analysis

```{r}
m_t2g <- msigdbr(species = "Mus musculus", category="H") %>%
  dplyr::select(gs_name, entrez_gene)
KO_genes_msigdb <- enricher(entrez_KO_genes, TERM2GENE=m_t2g)
WT_genes_msigdb <- enricher(entrez_WT_genes, TERM2GENE=m_t2g)

# Plot
dotplot_MSigDB_KO <- dotplot(KO_genes_msigdb, showCategory=10, font.size=12) + ggtitle("MSigDB Hallmark enrichment for KO-upregulated genes")
SaveFigure(res_path = res_path,
           plots = dotplot_MSigDB_KO,
           name = "dotplot_MSigDB_KO", 
           width = 15, 
           height = 8)

dotplot_MSigDB_WT <- dotplot(WT_genes_msigdb, showCategory=10, font.size=12) + ggtitle("MSigDB Hallmark enrichment for WT-upregulated genes")
SaveFigure(res_path = res_path,
           plots = dotplot_MSigDB_WT,
           name = "dotplot_MSigDB_WT", 
           width = 15, 
           height = 12)
```
## Identifying genes in a particular hallmark of interest

```{r}
# Import the msigdb hallmarks JSON file 
msigdb_hallmarks_json <- fromJSON(file="../miscellaneous/mh.all.v2023.2.Mm.json.txt")

# Obtain the gene symbols for the hallmark of interest. Replace HALLMARK_MYC_TARGETS_V1
# with the hallmark in question.

myc_gene_symbols <- msigdb_hallmarks_json$HALLMARK_MYC_TARGETS_V1$geneSymbols

# Filter the marker genes list to obtain information on the genes within the DE list
# that are present in the msigdb hallmarks file.

marker_genes_sig_condition %>%
  dplyr::filter(cluster == "KO") %>% 
  dplyr::filter(gene %in% myc_gene_symbols)
```
# Find DE genes between condition for macrophages

```{r}
# Subset for macrophages of interest
mbladder_m1_macro <- mbladder[,mbladder@meta.data$cell_type_annotations == "M1 Macrophages"]
mbladder_m2_macro <- mbladder[,mbladder@meta.data$cell_type_annotations == "M2 Macrophages"]

# Set identity for macrophages to be the KO/WT condition
Idents(mbladder_m1_macro) <- mbladder_m1_macro@meta.data$Condition
Idents(mbladder_m2_macro) <- mbladder_m2_macro@meta.data$Condition

# Find marker genes
marker_genes_m1_macro <- FindAllMarkers(mbladder_m1_macro, only.pos=TRUE)
marker_genes_m2_macro <- FindAllMarkers(mbladder_m2_macro, only.pos=TRUE)

# Filter for significant genes
marker_genes_m1_macro
marker_genes_sig_condition_m1Macro <- marker_genes_m1_macro %>%
  dplyr::filter(p_val_adj<0.05)
marker_genes_sig_condition_m2Macro <- marker_genes_m2_macro %>%
  dplyr::filter(p_val_adj<0.05)

marker_genes_sig_condition_m1Macro %>% write.csv(file=paste0(res_path, "marker_genes_m1Macro_onlyPos.csv"))
marker_genes_sig_condition_m2Macro %>% write.csv(file=paste0(res_path, "marker_genes_m2Macro_onlyPos.csv"))
```
# KEGG and MSigDB analysis of M1 macrophages between KO and WT

```{r}
# First, obtian the genes that are statistically significantly enrihed in KO vs WT
KO_enriched_genes_m1macro <- marker_genes_sig_condition_m1Macro %>%
  dplyr::filter(cluster == "KO") %>%
  pull()
marker_genes_sig_condition_m1Macro

WT_enriched_genes_m1macro <- marker_genes_sig_condition_m1Macro %>%
  dplyr::filter(cluster == "WT") %>%
  pull()

# Next, convert gene symbols to entrez ID by querying BioMart to convert symbols to Entrez IDs
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
# 
entrez_KO_genes_m1macro <- getBM(attributes = c('entrezgene_id'),
                     filters = 'mgi_symbol',
                     values = KO_enriched_genes_m1macro,
                     mart = ensembl) %>% pull()

entrez_WT_genes_m1macro <- getBM(attributes = c('entrezgene_id'),
                     filters = 'mgi_symbol',
                     values = WT_enriched_genes_m1macro,
                     mart = ensembl) %>% pull()
# Perform KEGG Overrepresentation analysis
KO_genes_kegg_m1macro <- enrichKEGG(gene = entrez_KO_genes_m1macro,
                 organism     = 'mmu',
                 pvalueCutoff = 0.05)

WT_genes_kegg_m1macro <- enrichKEGG(gene = entrez_WT_genes_m1macro,
                 organism     = 'mmu',
                 pvalueCutoff = 0.05)
# Plot
dotplot_KEGG_KO_m1macro <- dotplot(KO_genes_kegg_m1macro, showCategory=10, font.size=12) + ggtitle("KEGG Pathway enrichment for KO-upregulated genes in M1 macrophages")
dotplot_KEGG_KO_m1macro

dotplot_KEGG_WT_m1macro <- dotplot(WT_genes_kegg_m1macro, showCategory=10, font.size=12) + ggtitle("KEGG Pathway enrichment for WT-upregulated gene for M1 macrophages")
dotplot_KEGG_WT_m1macro

m_t2g <- msigdbr(species = "Mus musculus", category="H") %>%
  dplyr::select(gs_name, entrez_gene)
KO_genes_msigdb_m1macro <- enricher(entrez_KO_genes_m1macro, TERM2GENE=m_t2g)
WT_genes_msigdb_m1macro <- enricher(entrez_WT_genes_m1macro, TERM2GENE=m_t2g)

# Plot
dotplot_MSigDB_KO_m1macro <- dotplot(KO_genes_msigdb_m1macro, showCategory=10, font.size=12) + ggtitle("MSigDB Hallmark enrichment for KO-upregulated genes in M1 macrophages")
dotplot_MSigDB_KO_m1macro
SaveFigure(res_path = res_path,
           plots = dotplot_MSigDB_KO,
           name = "dotplot_MSigDB_KO_m1macro", 
           width = 15, 
           height = 8)

dotplot_MSigDB_WT_m1macro <- dotplot(WT_genes_msigdb_m1macro, showCategory=10, font.size=12) + ggtitle("MSigDB Hallmark enrichment for WT-upregulated genes")
dotplot_MSigDB_WT_m1macro
SaveFigure(res_path = res_path,
           plots = dotplot_MSigDB_WT_m1macro,
           name = "dotplot_MSigDB_WT_m1macro", 
           width = 15, 
           height = 12)
```
# KEGG and MSigDB analysis of M2 macrophages between KO and WT

```{r}
# First, obtian the genes that are statistically significantly enrihed in KO vs WT
KO_enriched_genes_m2macro <- marker_genes_sig_condition_m2Macro %>%
  dplyr::filter(cluster == "KO") %>%
  pull()


WT_enriched_genes_m2macro <- marker_genes_sig_condition_m2Macro %>%
  dplyr::filter(cluster == "WT") %>%
  pull()

# Next, convert gene symbols to entrez ID by querying BioMart to convert symbols to Entrez IDs
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
# 
entrez_KO_genes_m2macro <- getBM(attributes = c('entrezgene_id'),
                     filters = 'mgi_symbol',
                     values = KO_enriched_genes_m2macro,
                     mart = ensembl) %>% pull()

entrez_WT_genes_m2macro <- getBM(attributes = c('entrezgene_id'),
                     filters = 'mgi_symbol',
                     values = WT_enriched_genes_m2macro,
                     mart = ensembl) %>% pull()
# Perform KEGG Overrepresentation analysis
KO_genes_kegg_m2macro <- enrichKEGG(gene = entrez_KO_genes_m2macro,
                 organism     = 'mmu',
                 pvalueCutoff = 0.05)

WT_genes_kegg_m2macro <- enrichKEGG(gene = entrez_WT_genes_m2macro,
                 organism     = 'mmu',
                 pvalueCutoff = 0.05)
# Plot
dotplot_KEGG_KO_m2macro <- dotplot(KO_genes_kegg_m2macro, showCategory=10, font.size=12) + ggtitle("KEGG Pathway enrichment for KO-upregulated genes in M2 macrophages")
dotplot_KEGG_KO_m2macro

dotplot_KEGG_WT_m2macro <- dotplot(WT_genes_kegg_m2macro, showCategory=10, font.size=12) + ggtitle("KEGG Pathway enrichment for WT-upregulated gene for M1 macrophages")
dotplot_KEGG_WT_m2macro

m_t2g <- msigdbr(species = "Mus musculus", category="H") %>%
  dplyr::select(gs_name, entrez_gene)
KO_genes_msigdb_m2macro <- enricher(entrez_KO_genes_m2macro, TERM2GENE=m_t2g)
WT_genes_msigdb_m2macro <- enricher(entrez_WT_genes_m2macro, TERM2GENE=m_t2g)

# Plot
dotplot_MSigDB_KO_m2macro <- dotplot(KO_genes_msigdb_m2macro, showCategory=10, font.size=12) + ggtitle("MSigDB Hallmark enrichment for KO-upregulated genes in M2 macrophages")
dotplot_MSigDB_KO_m2macro
SaveFigure(res_path = res_path,
           plots = dotplot_MSigDB_KO_m2macro,
           name = "dotplot_MSigDB_KO_m2macro", 
           width = 15, 
           height = 8)

dotplot_MSigDB_WT_m2macro <- dotplot(WT_genes_msigdb_m2macro, showCategory=10, font.size=12) + ggtitle("MSigDB Hallmark enrichment for KO-downregulated genes in M2 macrophages")
SaveFigure(res_path = res_path,
           plots = dotplot_MSigDB_WT_m2macro,
           name = "dotplot_MSigDB_WT_m2macro", 
           width = 15, 
           height = 12)
```

