---
title: "02D_subclustering"
output: html_document
date: "2023-08-08"
---
```{r chunk_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, 
                      message=FALSE,
                      echo=TRUE,
                      include=TRUE)
```

# Overview

This 3rd analysis notebook showcases the steps involved in identifying a cluster
of interest and then subclustering it with a particular resolution using Seurat's
`FindSubCluster` command. This is followed by investigating the marker genes
for the identified subclusters, plotting them in the form of a heatmap,
and then renaming the clusters appropriately based on biology. 

Finally, we also conduct a differential abundance analysis test to determine
if the cells n the identified clusters differ between conditions (KO/WT).

# Load Libraries
```{r library}
library(Seurat)
library(dplyr)
library(tidyr)
library(ggplot2)
library(miloR)
library(SingleCellExperiment)
```

```{r globalVar}
# Loads in custom functions that I've written to assist with plotting/saving tasks
source("helper.R")

# Path to saving the results fo the analysis
res_path <- "../results/05_differential_abundance/"
if (!dir.exists(res_path)){
  
  dir.create(res_path)
}else{
  print("Results directory exists")
}
# Data path specifes where all the outputs from the parseBio split-pipe pipeline is stored.
data_path <- "../data/"
if (!dir.exists(data_path)){
  warning("Directory does not exist: ", data_path)
}else{
  print("Data directory exists")
}

# RDS path specifies where all the intermediate generated R objects will be saved.
rds_path <- "../rds/"
if (!dir.exists(rds_path)){
  warning("Directory does not exist: ", rds_path)
}else{
  print("RDS directory exists")
}
```


# Load annotated seurat object from 02.rmd
```{r load-qced-object}
# Import R object 
mbladder <- ReadObject(rds_path, "03_seuratMerged_annotated")
```

# Run Milo for DA analysis

```{r}
# First, convert seurat to sce object
mbladder_sce <- as.SingleCellExperiment(mbladder)

# Next, prepare Milo object
mbladder_sce_milo <- Milo(mbladder_sce)

# Build the KNN graph
traj_milo <- buildGraph(mbladder_sce_milo, k=30, d=30)

# Defining representative neighborhoods
traj_milo <- makeNhoods(traj_milo, prop = 0.1, k = 30, d=30, refined = TRUE)

# Counting cells in neighborhoods
traj_milo <- countCells(traj_milo, meta.data = data.frame(colData(traj_milo)), sample="sample_name")

# Differential abundance testing
## defining design matrix
traj_design <- data.frame(colData(traj_milo))[,c("sample_name", "Condition")]
traj_design <- distinct(traj_design)
rownames(traj_design) <- traj_design$sample_name
traj_design$Condition <- factor(traj_design$Condition, levels=c("WT", "KO"))

traj_milo <- calcNhoodDistance(traj_milo, d=30)
traj_milo <- buildNhoodGraph(traj_milo)

da_results <- testNhoods(traj_milo, design = ~ Condition, design.df = traj_design)
da_results <- annotateNhoods(traj_milo, da_results, coldata_col = "cell_type_annotations")

#da_results$macro_sub_mixed<- ifelse(da_results$macrophages_sub_fraction < 0.7, "Mixed", da_results$ident)
da_results %>%
  dplyr::filter(SpatialFDR < 0.1) %>%
  dplyr::arrange(logFC)
```

# Plotting Milo results

We observe that M1 macrophages appear to be more abundant in the KO condition, along 
with the inflammatory TAMs and other stem-like cells, whereas M2 macrophages 
are more abundant in the WT condition.

```{r}
top <- DimPlot(mbladder, reduction = "umap",  pt.size = 0.5,
        repel=TRUE, label=FALSE, label.box=TRUE, cols=stallion)
bottom <-  plotDAbeeswarm(da_results, group.by = "cell_type_annotations", alpha=0.1) + 
scater::plotUMAP(traj_milo, colour_by="Condition")  + plotNhoodGraphDA(traj_milo, da_results, alpha=0.05)
png(paste0(res_path, "results/milo_mbladder.png"), res=300, width = 50, height= 40 , units="cm")
cowplot::plot_grid(top, bottom,
          labels="AUTO", nrow = 2)
dev.off()

da_results
```
